---
title: "Thesis"
author: "Lillian Bradley"
date: "2025-03-15"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
\newpage

\newpage
# Libraries 
```{r}
library(ggplot2)
library(ggmosaic)
library(dplyr)
library(readr)
library(glmmTMB)
library(ggforce)
library(readr)
library(tidyr)
library(viridis)
library(cluster)
library(ggimage)
library(grid)
library(png)
library(xgboost)
library(caret)
library(Matrix)
library(broom)
library(purrr)
library(tidyverse)
library(raster)
library(vcd)
library(broom)
library(pROC)
library(randomForest)
```

\newpage
# Reading in Data
```{r}
#accessing data from file on desktop
base_dir <- "~/Desktop/Baseball/ncaa_trackman/"
all_files <- list.files(base_dir, recursive = TRUE, full.names = TRUE)

#player positioning data
player_pos_files <- grep("Robertson Field.*playerpositioning", all_files, value = TRUE)
player_pos_list <- lapply(player_pos_files, function(file) {
  df <- read_csv(file, col_types = cols(.default = "c")) 
  return(df)
})
player_pos_all <- bind_rows(player_pos_list)
head(player_pos_all, 2)

#pitch data
pitch_files <- setdiff(grep("Robertson Field", all_files, value = TRUE), player_pos_files)
pitch_list <- lapply(pitch_files, function(file) {
  df <- read_csv(file, col_types = cols(.default = "c"))  
  return(df)
})
pitch_all <- bind_rows(pitch_list)
head(pitch_all, 2)

#joining datasets 
columbia <- pitch_all %>%
  inner_join(player_pos_all, by = c("GameUID", "PlayID", "PitchUID", "PitcherTeam", "BatterTeam", "PitchCall", "PlayResult")) %>%
  #filtering data so examining just Columbia's defensive data 
  filter(PitcherTeam == "COL_LION") %>%
  dplyr::select(
    GameUID, PlayID, PitchUID, PAofInning, PitchofPA, Pitcher, PitcherId, PitcherThrows, PitcherTeam, 
    BatterSide, BatterTeam, Inning, Outs, Balls, Strikes, TaggedPitchType, 
    PitchCall, KorBB, TaggedHitType, PlayResult, OutsOnPlay, RunsScored, 
    DetectedShift, 
    `1B_PositionAtReleaseX`, `1B_PositionAtReleaseZ`, 
    `2B_PositionAtReleaseX`, `2B_PositionAtReleaseZ`, 
    `3B_PositionAtReleaseX`, `3B_PositionAtReleaseZ`, 
    `SS_PositionAtReleaseX`, `SS_PositionAtReleaseZ`, 
    `LF_PositionAtReleaseX`, `LF_PositionAtReleaseZ`, 
    `CF_PositionAtReleaseX`, `CF_PositionAtReleaseZ`, 
    `RF_PositionAtReleaseX`, `RF_PositionAtReleaseZ`
  )

head(columbia,3)

```

```{r}
#number of games
length(unique(columbia$GameUID))

```


\newpage
# Exporatory Data Analysis (EDA)
```{r}
#comparing pitcher and batter handedness

#mosaic plot
ggplot(columbia) +
  geom_mosaic(aes(x = product(PitcherThrows, BatterSide), fill = BatterSide)) +
  scale_fill_manual(values = c("Left" = "#012169", "Right" = "#6CACE4")) +  
  labs(title = "Relationship between Pitcher and Batter Hand - Columbia Baseball @ Home",
       x = "Pitcher Handedness",
       y = "Proportion",
       fill = "Batter Side") +
  theme_minimal()



#tacked bar chart
ggplot(columbia, aes(x = PitcherThrows, fill = BatterSide)) +
  geom_bar(position = "fill") +  
  scale_fill_manual(values = c("Left" = "#012169", "Right" = "#6CACE4")) + 
  labs(title = "Pitcher Hand vs. Batter Hand - Columbia Baseball 2022-2024",
       x = "Pitcher Handedness",
       y = "Proportion",
       fill = "Batter Side") +
  theme_minimal()

```

```{r}
#long formatting for easier plotting of positions 
col_long <- columbia %>%
  dplyr::select(matches("PositionAtReleaseX|PositionAtReleaseZ")) %>%
  pivot_longer(cols = everything(), names_to = "Position", values_to = "Coordinate") %>%
  separate(Position, into = c("Player", "Axis"), sep = "PositionAtRelease") %>%
  pivot_wider(names_from = Axis, values_from = Coordinate)

head(col_long)
col_long <- col_long %>%
  unnest(cols = c(X, Z)) %>%
  mutate(X = as.numeric(X), Z = as.numeric(Z)) %>%
  drop_na()

head(col_long)
```
Density heat-map showing seven distinct defensive positions. 


```{r}
#loading the field image
robertson_field_path <- "~/Desktop/Thesis/columbia_field.png"
robertson_field <- readPNG(robertson_field_path)


col_long <- na.omit(col_long)

#adjusting plot limits for better visualization
x_min <- min(col_long$Z, na.rm = TRUE) - 10
x_max <- max(col_long$Z, na.rm = TRUE) + 10
y_min <- min(col_long$X, na.rm = TRUE) - 10
y_max <- max(col_long$X, na.rm = TRUE) + 10

#setting offset value for alignment 
offset_x <- 30  

col_long_shifted <- col_long
col_long_shifted$Z <- col_long_shifted$Z + offset_x  

#plotting heatmap of positions overlaying image of field 
ggplot(col_long_shifted, aes(x = Z, y = X)) + 
  annotation_custom(
    rasterGrob(robertson_field, width = unit(1.5, "npc"), height = unit(1.5, "npc")),
    xmin = x_min, xmax = x_max, ymin = y_min, ymax = y_max
  ) + 
  geom_bin2d(bins = 30, alpha = 0.7) + 
  scale_fill_viridis_c(option = "plasma") + 
  theme_minimal() +
  labs(
    title = "Heatmap of Position Players on Robertson Field (Columbia University)",
    x = "Z Position (Horizontal)", 
    y = "X Position (Vertical)"
  ) + 
  coord_fixed(ratio = 1)  
```

heatmap with position specific colors
```{r}
col_long_shifted$Position <- factor(col_long_shifted$Player, levels = c("1B_", "2B_", "3B_",
                                                        "SS_", "LF_", "CF_", "RF_"))

#specifying colors for each position
pos_colors <- c("1B_" = "pink", "2B_" = "blue", "3B_" = "yellow", "SS_" = "green", 
                "LF_" = "red", "CF_" = "purple", "RF_" = "orange")

#plotting heatmap of positions overlaying image of field 
ggplot(col_long_shifted, aes(x = Z, y = X)) + 
  annotation_custom(
    rasterGrob(robertson_field, width = unit(1.5, "npc"), height = unit(1.5, "npc")),
    xmin = x_min, xmax = x_max, ymin = y_min, ymax = y_max
  ) + 
  geom_bin2d(bins = 30, alpha = 0.7, aes(fill = Position)) +  
  scale_fill_manual(values = pos_colors) +  
  theme_minimal() +
  labs(
    title = "Colorcoded Heatmap of Position Players on Robertson Field (Columbia University)",
    x = "Z Position (Horizontal)", 
    y = "X Position (Vertical)",
    fill = "Player Position"  
  ) + 
  coord_fixed(ratio = 1) 
```

examining shift 
```{r}
shift_play_result_summary <- columbia %>%
  group_by(DetectedShift, PlayResult) %>%
  tally() %>%
  arrange(DetectedShift, PlayResult)

#summary table
print(shift_play_result_summary)

#barplot
ggplot(shift_play_result_summary, aes(x = DetectedShift, y = n, fill = PlayResult)) +
  geom_bar(stat = "identity", position = "stack") +  # Stack the bars
  theme_minimal() +
  labs(
    title = "Play Results by Defensive Shift",
    x = "Detected Shift",
    y = "Count of Play Results",
    fill = "Play Result"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#percent of time out is result from shift
shift_data <- columbia %>%
  filter(DetectedShift %in% c("LeftInfieldShift", "RightInfieldShift"))

# Calculate the percentage of "Out" PlayResults when a shift is detected
out_percentage_when_shifted <- shift_data %>%
  summarise(percentage_out = mean(PlayResult == "Out") * 100)

print(out_percentage_when_shifted)

#percent of pitches there is a shift
total_shifts_percentage <- columbia %>%
  summarise(percentage_with_shift = mean(DetectedShift %in% c("LeftInfieldShift", "RightInfieldShift")) * 100)

print(total_shifts_percentage)

#percent of times theres a hit against shift 
hits <- c("Single", "Double", "HomeRun", "Triple")

filtered_data <- columbia %>%
  filter(PlayResult %in% hits)

play_result_percentages <- filtered_data %>%
  group_by(PlayResult) %>%
  summarise(count = n()) %>%
  mutate(percentage = (count / nrow(columbia)) * 100)  

print(play_result_percentages)

ggplot(play_result_percentages, aes(x = PlayResult, y = percentage, fill = PlayResult)) +
  geom_bar(stat = "identity") +  # Use the actual values (percentage) for the bar heights
  scale_fill_manual(values = c("Single" = "#003865", "Double" = "#9BCBEB", "HomeRun" = "#BD3039", "Triple" = "#DF4601")) + 
  theme_minimal() +
  labs(
    title = "Hit Type Distrubtion With Defensive Shift Implemented",
    x = "Play Result",
    y = "Percentage (%)"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_text(aes(label = paste0(count, " (", round(percentage, 1), "%)")), 
            position = position_stack(vjust = 0.5), size = 4, color = "#FFFFFF")

```

breaking down heatmaps by handedness (batter and pitcher)

preparing data
```{r}
col_expanded <- columbia %>% 
  dplyr::select(PitchUID, PitcherThrows, BatterSide, PlayResult, DetectedShift, matches("PositionAtReleaseX|PositionAtReleaseZ"))

col_expanded <- col_expanded %>%
  mutate(across(contains("PositionAtRelease"), as.numeric)) %>%
  drop_na()

head(col_expanded)

col_expanded_long <- col_expanded %>%
  pivot_longer(cols = matches("PositionAtReleaseX|PositionAtReleaseZ"), names_to = "Position", values_to = "Coordinate") %>%
  separate(Position, into = c("Player", "Axis"), sep = "PositionAtRelease") %>%
  pivot_wider(names_from = Axis, values_from = Coordinate)

col_expanded_long <- col_expanded_long %>%
  unnest(cols = c(X, Z)) %>%
  mutate(X = as.numeric(X), Z = as.numeric(Z)) %>%
  drop_na()

head(col_expanded_long)

```
plotting heatmaps
```{r}
#removing undefined
col_expanded_long <- col_expanded_long %>%
  filter(PitcherThrows %in% c("Left", "Right"), BatterSide %in% c("Left", "Right"))

#adjusting plot
x_min <- min(col_expanded_long$Z, na.rm = TRUE) - 10
x_max <- max(col_expanded_long$Z, na.rm = TRUE) + 10
y_min <- min(col_expanded_long$X, na.rm = TRUE) - 10
y_max <- max(col_expanded_long$X, na.rm = TRUE) + 10


offset_x <- 30

col_expanded_long$Z <- col_expanded_long$Z + offset_x

#plotting based on handedness
ggplot(col_expanded_long, aes(x = Z, y = X)) + 
  annotation_custom(
    rasterGrob(robertson_field, width = unit(1.75, "npc"), height = unit(1.75, "npc")), 
    xmin = x_min, xmax = x_max, ymin = y_min, ymax = y_max
  ) + 
  geom_bin2d(bins = 30, alpha = 0.7) +  
  scale_fill_viridis_c(option = "plasma") +  
  facet_grid(PitcherThrows ~ BatterSide) +  
  theme_minimal() + 
  labs(
    title = "Heatmap of Defensive Positioning on Robertson Field (Columbia University)",
    x = "Batter Handedness (Horizontal)", 
    y = "Pitcher Handedness (Vertical)",
    subtitle = "Pitcher Handedness vs Batter Handedness",
    caption = "Each plot represents the distribution of defensive positioning by handedness."
  ) +  
  theme(
    strip.text = element_text(size = 12, face = "bold"),  
    plot.title = element_text(size = 16, face = "bold"), 
    plot.subtitle = element_text(size = 12, face = "italic") 
  ) + 
  coord_fixed(ratio = 1)  
```


color coded for positions heatmap with handedness 
```{r}
#plotting based on handedness
ggplot(col_expanded_long, aes(x = Z, y = X)) + 
  annotation_custom(
    rasterGrob(robertson_field, width = unit(1.75, "npc"), height = unit(1.75, "npc")), 
    xmin = x_min, xmax = x_max, ymin = y_min, ymax = y_max
  ) + 
  #color specific for positions
  geom_bin2d(bins = 30, alpha = 0.7, aes(fill = Player)) +  
  scale_fill_manual(values = pos_colors) +  
  facet_grid(PitcherThrows ~ BatterSide) +  
  theme_minimal() + 
  labs(
    title = "Heatmap of Defensive Positioning on Robertson Field (Columbia University)",
    x = "Batter Handedness (Horizontal)", 
    y = "Pitcher Handedness (Vertical)",
    subtitle = "Pitcher Handedness vs Batter Handedness",
    caption = "Each plot represents the distribution of defensive positioning by handedness."
  ) + 
  theme(
    strip.text = element_text(size = 12, face = "bold"),  
    plot.title = element_text(size = 16, face = "bold"),  
    plot.subtitle = element_text(size = 12, face = "italic")  
  ) + 
  coord_fixed(ratio = 1)  
```

which positions cover the greatest area 
```{r}
area_per_position <- col_expanded_long %>%
  group_by(Player) %>%
  summarise(
    min_X = min(X, na.rm = TRUE),
    max_X = max(X, na.rm = TRUE),
    min_Z = min(Z, na.rm = TRUE),
    max_Z = max(Z, na.rm = TRUE),
    area = (max_X - min_X) * (max_Z - min_Z)
  ) %>%
  arrange(desc(area))

print(area_per_position)

#visualizing total area 
ggplot(area_per_position, aes(x = reorder(Player, area), y = area)) +
  geom_bar(stat = "identity", fill = "#9BCBEB") +
  geom_text(aes(label = round(area, 1)), vjust = -0.5, color = "#003865", size = 4, hjust = 1.1) +
  theme_minimal() +
  labs(
    title = "Area Covered by Each Position on the Field",
    x = "Player Position",
    y = "Area Covered (Square Units)"
  ) +
  coord_flip()

```

# Models

basic model 
```{r}
#checking how many pitches there are total 
cat("\nNumber of pitches:", length(unique(col_expanded$PitchUID)))

#checking how many pitches there are after removing undefined 
cat("\nNumber of pitches:", 
    length(unique(col_expanded$PitchUID[col_expanded$PlayResult == "Undefined"])))

col_expanded <- col_expanded %>%
    filter(PlayResult != "Undefined")

#checking how many pitches there are after removing Undefined results 
cat("\nNumber of pitches:", length(unique(col_expanded$PitchUID)))

#checking how many pitches resulted in an out
cat("\nNumber of pitches:", length(unique(col_expanded$PitchUID[col_expanded$PlayResult == "Out"])))

#converting play result to binary, 1 = out, 0 = not out
col_expanded$PlayResult <- ifelse(col_expanded$PlayResult == "Out", 1, 0)

#converting categorical variables to factors
col_expanded <- col_expanded %>%
  mutate(
    BatterSide = factor(BatterSide),
    PitcherThrows = factor(PitcherThrows), 
    DetectedShift = factor(DetectedShift)
  )
#printing what data looks like 
head(col_expanded, 7)
```

splitting data into training and testing
```{r}
set.seed(22)
train_indices <- sample(1:nrow(col_expanded), size = 0.8 * nrow(col_expanded))
train_data <- col_expanded[train_indices, ]
test_data <- col_expanded[-train_indices, ]

dim(train_data)
dim(test_data)
```

linear regression models - one for x and one for z
```{r}
#logistic regression model to predict PlayResult 
log_model <- glm(PlayResult ~ 
                   `1B_PositionAtReleaseX` + `1B_PositionAtReleaseZ` +
                   `2B_PositionAtReleaseX` + `2B_PositionAtReleaseZ` + 
                   `3B_PositionAtReleaseX` + `3B_PositionAtReleaseZ` + 
                   `SS_PositionAtReleaseX` + `SS_PositionAtReleaseZ` + 
                   `LF_PositionAtReleaseX` + `LF_PositionAtReleaseZ` + 
                   `CF_PositionAtReleaseX` + `CF_PositionAtReleaseZ` + 
                   `RF_PositionAtReleaseX` + `RF_PositionAtReleaseZ` + 
                   BatterSide + PitcherThrows + DetectedShift, 
                 data = train_data, family = "binomial")

summary(log_model)

#predictions on test data
pred_probs <- predict(log_model, newdata = test_data, type = "response")
predictions <- ifelse(pred_probs > 0.5, 1, 0)

#confusion matrix 
conf_matrix <- table(Predicted = predictions, Actual = test_data$PlayResult)
print(conf_matrix)

#calculate accuracy
accuracy <- sum(predictions == test_data$PlayResult) / length(predictions)
cat("\nAccuracy: ", accuracy)

#roc curve and auc
roc_curve <- roc(test_data$PlayResult, pred_probs)
plot(roc_curve, col = "#003865", lwd = 2, main = "ROC Curve for Predicting Outs in Baseball",
     xlab = "False Positive Rate", ylab = "True Positive Rate")
cat("\nAUC: ", auc(roc_curve))

```

# extreme gradient boost model 
```{r}
set.seed(42)

#preparing the fearures  
train_data_X <- train_data %>%
  dplyr::select(-PlayResult) %>%
  mutate(across(everything(), as.numeric))
test_data_X <- test_data %>%
  dplyr::select(-PlayResult) %>%
  mutate(across(everything(), as.numeric))

#converting data into DMatrix format
dtrain <- xgb.DMatrix(data = as.matrix(train_data_X), label = train_data$PlayResult)
dtest <- xgb.DMatrix(data = as.matrix(test_data_X), label = test_data$PlayResult)
```

training model 
```{r}
#hyperparams
params <- list(
  objective = "reg:squarederror", 
  booster = "gbtree",            
  eta = 0.1,                       
  max_depth = 6,                  
  nthread = 2                      
)

#cross-validation to find optimal number of rounds 
cv_results <- xgb.cv(
  params = params,
  data = dtrain,
  nrounds = 100,
  nfold = 5,
  stratified = TRUE,
  early_stopping_rounds = 10,
  maximize = FALSE,
  verbose = 1
)

best_nrounds <- cv_results$best_iteration
cat("Best number of rounds:", best_nrounds, "\n")

#training final model
model <- xgb.train(
  params = params,
  data = dtrain,
  nrounds = best_nrounds,
  verbose = 1
)
```

prediction models on test data 
```{r}
#predictions on test set
pred_probs <- predict(model, dtest)

#converting probabilities to binary predictions (using 0.5 (random) as threshold)
predictions <- ifelse(pred_probs > 0.5, 1, 0)

#confusion matrix
conf_matrix_xg <- table(Predicted = predictions, Actual = test_data$PlayResult)
print(conf_matrix_xg)

#accuracy
accuracy_xg <- sum(predictions == test_data$PlayResult) / length(predictions)
cat("\nAccuracy: ", accuracy_xg, "\n")

#ROC curve and AUC
roc_curve_xg <- roc(test_data$PlayResult, pred_probs)
plot(roc_curve_xg, col = "#003865", lwd = 2, main = "ROC Curve for Predicting Outs in Baseball",
     xlab = "False Positive Rate", ylab = "True Positive Rate")
cat("\nAUC: ", auc(roc_curve_xg))
```

# Random Forest Model
```{r}
#setting PlayResult as factor
train_data$PlayResult <- as.factor(train_data$PlayResult)
test_data$PlayResult <- as.factor(test_data$PlayResult)

#setting up training control for cross validation
train_control <- trainControl(method = "cv", number = 5, search = "grid")

#random forest model training 
rf_grid <- expand.grid(mtry = c(3, 5, 7))  

head(train_data, 2)

rf_model <- train(PlayResult ~ `1B_PositionAtReleaseX` + `1B_PositionAtReleaseZ` + 
                  `2B_PositionAtReleaseX` + `2B_PositionAtReleaseZ` + 
                  `3B_PositionAtReleaseX` + `3B_PositionAtReleaseZ` + 
                  `SS_PositionAtReleaseX` + `SS_PositionAtReleaseZ` + 
                  `LF_PositionAtReleaseX` + `LF_PositionAtReleaseZ` + 
                  `CF_PositionAtReleaseX` + `CF_PositionAtReleaseZ` + 
                  `RF_PositionAtReleaseX` + `RF_PositionAtReleaseZ` +
                  BatterSide + PitcherThrows + DetectedShift, 
                data = train_data, 
                method = "rf", 
                trControl = train_control, 
                tuneGrid = rf_grid, 
                importance = TRUE,
                family = "binomial")

#best parameters 
print(rf_model$bestTune)
summary(rf_model)

```

predictions on test set
```{r}
predictions <- predict(rf_model, newdata = test_data)
pred_probs_rf <- predict(rf_model, newdata = test_data, type = "prob")[, 2]

#confusion matrix
conf_matrix <- confusionMatrix(predictions, test_data$PlayResult)
print(conf_matrix)

#accuracy
accuracy_rf <- sum(predictions == test_data$PlayResult) / length(predictions)
cat("\nAccuracy: ", accuracy_rf)

#feature importance plot
varImpPlot(rf_model$finalModel)

#roc and auc
roc_curve_rf <- roc(test_data$PlayResult, pred_probs_rf)
plot(roc_curve_rf, col = "#003865", lwd = 2, main = "ROC Curve for Predicting Outs in Baseball Using a Random Forest Model",
     xlab = "False Positive Rate", ylab = "True Positive Rate")
cat("\nAUC: ", auc(roc_curve))
```

# comparing the models
```{r}
#ROC curves for all models
plot(roc_curve, col = "#003865", lwd = 2, main = "ROC Curve Comparison", xlab = "False Positive Rate", ylab = "True Positive Rate")
plot(roc_curve_rf, col = "#9BCBEB", lwd = 2, add = TRUE)
plot(roc_curve_xg, col = "#bd3039", lwd = 2, add = TRUE)
legend("bottomright", legend = c("Logistic Regression", "Random Forest", "XGBoost"), col = c("#003865", "#9BCBEB", "#bd3039"), lwd = 2)

```

# final model - xgboost 

## preparing the data 
```{r}
train_data$PlayResult <- as.factor(train_data$PlayResult)
test_data$PlayResult <- as.factor(test_data$PlayResult)

#adding pitcher batter handedness intercation
train_data$BatterPitcherInteraction <- interaction(train_data$BatterSide, train_data$PitcherThrows)
test_data$BatterPitcherInteraction <- interaction(test_data$BatterSide, test_data$PitcherThrows)

train_data$PlayResult <- as.numeric(train_data$PlayResult) - 1
test_data$PlayResult <- as.numeric(test_data$PlayResult) - 1

train_data_X <- train_data %>%
  dplyr::select(-PlayResult) %>%
  mutate(across(everything(), as.numeric))
test_data_X <- test_data %>%
  dplyr::select(-PlayResult) %>%
  mutate(across(everything(), as.numeric))

#convert to DMatrix
dtrain <- xgb.DMatrix(data = as.matrix(train_data_X), label = train_data$PlayResult)
dtest <- xgb.DMatrix(data = as.matrix(test_data_X), label = test_data$PlayResult)

```

tuning hyperparameters 
```{r}
#using grid search to tune hyperparameters 
params_grid <- expand.grid(
  eta = c(0.01, 0.05, 0.1, 0.2),
  max_depth = c(3, 5, 6, 10),
  subsample = c(0.6, 0.8, 1),
  colsample_bytree = c(0.6, 0.8, 1),
  gamma = c(0, 0.1, 0.2)
)

#grid search
for (i in 1:nrow(params_grid)) {
  params <- list(
    objective = "binary:logistic",  
    booster = "gbtree",
    eta = params_grid$eta[i],
    max_depth = params_grid$max_depth[i],
    subsample = params_grid$subsample[i],
    colsample_bytree = params_grid$colsample_bytree[i],
    gamma = params_grid$gamma[i],
    nthread = 2
  )
  
  #cross validation 
  cv_results <- xgb.cv(
    params = params,
    data = dtrain,
    nrounds = 100,
    nfold = 5,
    early_stopping_rounds = 10,
    verbose = 1,
    maximize = FALSE
  )
  
  best_nrounds <- cv_results$best_iteration
  cat("\nBest number of rounds for params set", i, ":", best_nrounds, "\n")
}

```

## training final model 
```{r}
# Using the best parameters from the grid search
best_params <- list(
  objective = "binary:logistic",  
  booster = "gbtree",
  eta = 0.1,  
  max_depth = 5,  
  subsample = 0.8,
  colsample_bytree = 1,
  gamma = 0.1,
  nthread = 2
)

# Train the final model using the best nrounds
final_model <- xgb.train(
  params = best_params,
  data = dtrain,
  nrounds = best_nrounds,
  verbose = 1
)

```

# visualizing results and such
```{r}
#predictions on test set
pred_probs_final <- predict(final_model, dtest)

#converting probabilities to binary predictions (using 0.5 (random) as threshold)
predictions_final <- ifelse(pred_probs_final > 0.5, 1, 0)

#confusion matrix
conf_matrix_xg_final <- table(Predicted = predictions_final, Actual = test_data$PlayResult)
print(conf_matrix_xg_final)

#accuracy
accuracy_xg_final <- sum(predictions == test_data$PlayResult) / length(predictions_final)
cat("\nAccuracy: ", accuracy_xg_final, "\n")

#ROC curve and AUC
roc_curve_xg <- roc(test_data$PlayResult, pred_probs_final)
plot(roc_curve_xg, col = "#003865", lwd = 2, main = "ROC Curve for Predicting Outs in Baseball",
     xlab = "False Positive Rate", ylab = "True Positive Rate")
cat("\nAUC: ", auc(roc_curve_xg))
```

# Below Here is the Code I am still wokring on for optimization 
# finding optimal locations for each player



```{r}
#recreate the BatterPitcherInteraction column as a string combination of BatterSide and PitcherThrows
test_data$BatterPitcherInteraction <- paste(test_data$BatterSide, test_data$PitcherThrows, sep = ".")

find_optimal_positions <- function(batter_side, pitcher_throws) {
  interaction_label <- paste(batter_side, pitcher_throws, sep = ".")
  
  #filtering data based on batter and pitcher handedness
  test_data_filtered <- test_data[test_data$BatterPitcherInteraction == interaction_label, ]
  
  #checking ig filtered data has rows
  if (nrow(test_data_filtered) == 0) {
    cat("No data found for Batter:", batter_side, "Pitcher:", pitcher_throws, "\n")
    return(NULL)  
  }

  #baseline positions
  positions <- c(92.64412, 50.53915,  #1B
                 146.4060, 33.66025,  #2B
                 108.06077, -90.91847,  #3B
                 139.1949, -51.55803,  #SS
                 256.0358, -131.6172,  #LF
                 305.4376, 7.71895,  #CF
                 262.6032, 123.60749)  #RF
  
  #objective function for optimization
  optimize_positions <- function(positions) {
    test_data_optimized <- test_data_filtered
    
    #update player positions with the current ones
    test_data_optimized$X1B <- positions[1]
    test_data_optimized$Z1B <- positions[2]
    test_data_optimized$X2B <- positions[3]
    test_data_optimized$Z2B <- positions[4]
    test_data_optimized$X3B <- positions[5]
    test_data_optimized$Z3B <- positions[6]
    test_data_optimized$XSS <- positions[7]
    test_data_optimized$ZSS <- positions[8]
    test_data_optimized$XLF <- positions[9]
    test_data_optimized$ZLF <- positions[10]
    test_data_optimized$XCF <- positions[11]
    test_data_optimized$ZCF <- positions[12]
    test_data_optimized$XRF <- positions[13]
    test_data_optimized$ZRF <- positions[14]
    
    test_data_optimized <- test_data_optimized[, colnames(train_data)]
  
    test_data_optimized[] <- lapply(test_data_optimized, function(x) {
      if (is.factor(x)) return(as.numeric(as.character(x))) 
      if (is.logical(x)) return(as.numeric(x))  # TRUE -> 1, FALSE -> 0
      return(x)
    })
    
    #DMatrix format
    dtest_optimized <- xgb.DMatrix(data = as.matrix(test_data_optimized))
    
    #predictions from the trained model
    pred_probs_optimized <- predict(final_model, dtest_optimized)
    
    #total outs (predictions > 0.5)
    total_outs <- sum(pred_probs_optimized > 0.5)
    
    return(-total_outs)  # Negative for minimization
  }

  #initial positions vector
  initial_positions <- positions
  
  #apply optimization usingSimulated Annealing
  result <- optim(
    par = initial_positions,              
    fn = optimize_positions,              
    method = "BFGS",                       
    control = list(fnscale = -1)          
  )
  
  #optimal positions
  optimal_positions <- result$par
  
  cat("Optimal Positions for Batter:", batter_side, "Pitcher:", pitcher_throws, "\n")
  cat("1B (X, Z):", optimal_positions[1], optimal_positions[2], "\n")
  cat("2B (X, Z):", optimal_positions[3], optimal_positions[4], "\n")
  cat("3B (X, Z):", optimal_positions[5], optimal_positions[6], "\n")
  cat("SS (X, Z):", optimal_positions[7], optimal_positions[8], "\n")
  cat("LF (X, Z):", optimal_positions[9], optimal_positions[10], "\n")
  cat("CF (X, Z):", optimal_positions[11], optimal_positions[12], "\n")
  cat("RF (X, Z):", optimal_positions[13], optimal_positions[14], "\n")
  
  return(optimal_positions)
}

#Right-handed batter vs Right-handed pitcher
optimal_positions_right_right <- find_optimal_positions("Right", "Right")
#Left-handed batter vs Right-handed pitcher
optimal_positions_left_right <- find_optimal_positions("Left", "Right")
#Right-handed batter vs Left-handed pitcher
optimal_positions_right_left <- find_optimal_positions("Right", "Left")
#Left-handed batter vs Left-handed pitcher
optimal_positions_left_left <- find_optimal_positions("Left", "Left")


```

